Revealing transient strain in geodetic data with Gaussian process regression

### 2017-11-08 RevisedSubmission/hines_hetland_2017_gji.tex
did more

@@

The aforementioned studies are concerned with estimating secular
strain rates. In recent years the Southern California Earthquake
Center (SCEC) community has shown interest in developing methods for
detecting transient strain. SCEC supported a transient detection
exercise \citep{Lohman2013}, where several research groups tested
their methods for detecting transient geophysical signal with a
synthetic GNSS dataset. Among the methods tested were the Network
Strain Filter (NSF) \citep{Ohtani2010} and the Network Inversion
Filter (NIF) \citep{Segall1997}. The NSF uses a wavelet
parameterization to describe the spatial component of geophysical
signal. The NIF, which is intended for imaging slow fault slip from
geodetic data, uses the elastic dislocation Green's functions from
\citet{Okada1992}. For the NSF and NIF, the time dependence of the
geophysical signal is modeled as integrated Brownian motion. The
method described in \citet{Holt2013} was also tested in the SCEC
transient detection exercise, which calculates strain rates using a
bi-cubic spatial parameterization of displacements between time
epochs. \citet{Holt2013} defined a detection threshold based on the
strain rate magnitude, and below we demonstrate that this is indeed an
effective criterion for identifying geophysical signal. For the same
reasons descibed above, the transient deformation and corresponding
uncertainties estimated by these methods can be biased by the chosen
spatial parameterization. It is then difficult to distinguish signal
from noise with these methods, which limits their utility for
transient detection.

@@

Another significant source of noise in GNSS data is common mode error
\citep[e.g.,][]{Wdowinski1997,Dong2006}, which is noise that is highly
spatially correlated. When not accounted for, common mode error
manifests as spatially uniform undulations in estimated transient
displacements. However, estimated transient strain rates are
insensitive to common mode error, and so we do not need to include
common mode error in our noise model.

@@

%% DISCUSSION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{sec:Discussion}
Our results demonstrate that GPR is an effective tool for estimating
transient strain from GNSS data, which can then be used to detect
geophysical processes. One may argue that geophysical signal can also
be detected by merely inspecting the GNSS displacement time series.
Indeed, the SSEs identified in Figure \ref{fig:StrainMag} do produce
visible displacements in the GNSS data. However, the GNSS data also
contains outliers and non-tectonic deformation that is localized to
individual stations. In contrast, our estimates of transient strain
only identify features that are sufficiently spatially and temporally
coherent, based on our chosen prior model. Furthermore, our estimates
of transient strain are insensitive to common mode noise, which is
highly spatially correlated noise resulting from factors such as
reference frame error. Common mode noise can obscure geophysical
signal in the GNSS data, but it gets canceled out when computing the
transient strain. Lastly, our estimates of transient strain rates are
a spatial and temporal derivative of displacements, and thus any
geophysical signal in the transient strain rates tends to be more
pronounced than in the GNSS data. For these reasons, we argue that
transient strain rates estimated with the method described in Section
\ref{sec:Method} can illuminate geophysical signal that may not be
discernible from the noise in the GNSS displacement data.

@@

There is potential for a more thorough analysis of the spatio-temporal
noise in GNSS data, $\eta$, than what was performed in Section
\ref{sec:NoiseModel}. We did not explore the spatial covariance of
$\eta$, which would describe common mode noise. We are able to ignore
common mode error in this study; however, for other geophysical
studies based on GNSS data, such as fault slip inversions, it may be
necessary to incorporate a spatially covarying noise model
\citep[e.g.,][]{Miyazaki2003}. We can also improve upon the seasonal
model used in this study, which consists of four spatially
uncorrelated sinusoids for each station. We did not explore the
spatial covariance of seasonal deformation or the temporal roughness
(i.e., the number of sinusoids needed to describe the observations).
The periodic Gaussian process \citep{Mackay1998} is an alternative
model for seasonal deformation and is well suited for exploring the
roughness of seasonal deformation.  The periodic Gaussian process has
zero mean and the covariance function
\begin{equation}\label{eq:Periodic}
T(t,t') = \phi^2 \exp\left(\frac{-\sin(\pi|t - t'|)^2}{2\tau^2}\right).
\end{equation}
Realizations have annual periodicity and the roughness is controlled
by $\tau$. Decreasing $\tau$ has the same effect as including higher
frequency sinusoids in the seasonal model. The optimal value for
$\tau$ can be found with the REML method as described in Section
\ref{sec:NoiseModel}.

The transient strain rates estimated in this study are constrained by
about seven years of daily displacement observations from 94 GNSS
stations. It can be computationally intensive to evaluate eqs.
(\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) for a dataset
with this size. We significantly reduce the amount of memory needed to
estimate transient strain rates by describing the temporal covariance
of displacements with a compact Wendland covariance function. Using a
compact covariance function for our prior turns eqs.
(\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) into sparse
systems of equations, which we then solve with CHOLMOD. CHOLMOD is
designed for solving sparse, positive definite systems of equations.
The matrix being inverted in eqs. (\ref{eq:PosteriorMean3}) and
(\ref{eq:PosteriorCov3}) is not positive definite; however, we can use
another partitioned matrix inversion identity from \citet{Press2007}
to partition it into positive definite submatrices to be inverted.
Even when using a compact covariance function, it may still be
necessary to reduce the computational burden by dividing the data into
subsets and evaluating transient strain rates for each subset.

>>>>

% focus on subjective vs object rather than "we have uncertainties"

@@

The aforementioned studies are concerned with estimating secular
strain rates. In recent years the Southern California Earthquake
Center (SCEC) community has shown interest in developing methods for
detecting transient strain. SCEC supported a transient detection
exercise \citep{Lohman2013}, where several research groups tested
their methods for detecting transient geophysical signal with a
synthetic GNSS dataset. Among the methods tested were the Network
Strain Filter (NSF) \citep{Ohtani2010} and the Network Inversion
Filter (NIF) \citep{Segall1997}. The NSF uses a wavelet
parameterization to describe the spatial component of geophysical
signal. The NIF, which is intended for imaging slow fault slip from
geodetic data, uses the elastic dislocation Green's functions from
\citet{Okada1992}. For the NSF and NIF, the time dependence of the
geophysical signal is modeled as integrated Brownian motion. The
method described in \citet{Holt2013} was also tested in the SCEC
transient detection exercise, which calculates strain rates using a
bi-cubic spatial parameterization of displacements between time
epochs. \citet{Holt2013} defined a detection threshold based on the
strain rate magnitude, and we demonstrate that this is indeed an
effective criterion for identifying geophysical signal. For the same
reasons descibed above, the transient deformation and corresponding
uncertainties estimated by these methods can be biased by the chosen
spatial parameterization. It is then difficult to distinguish signal
from noise with these methods, which limits their utility for
transient detection.

@@

Another significant source of noise in GNSS data is common mode error
\citep[e.g.,][]{Wdowinski1997,Dong2006}, which is noise that is highly
spatially correlated. When not accounted for, common mode error
manifests as spatially uniform undulations in our estimated transient
displacements. These undulations have no effect on the transient
strain rates, and so we do not need to include common mode error in
our noise model.

@@

%% DISCUSSION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{sec:Discussion}

% issues and potential improvements on noise

Our results demonstrate that GPR is an effective tool for estimating
transient strain rates and detecting geophysical processes from GNSS
data. However, there are some aspects of our method that warrant
further research. For example, there is potential for a more thorough
analysis of the spatio-temporal noise in the GNSS data than what was
performed in Section \ref{sec:NoiseModel}. Specifically, we did not
explore spatially correlated noise, such as from common mode error. We
have assumed that any spatially correlated noise has a sufficiently
long wavelength that it has a negligible effect on transient strain
rates. We can also improve upon the seasonal noise model used in this
study, which consists of four spatially uncorrelated sinusoids for
each station. We did not explore the spatial covariance of seasonal
deformation or the temporal roughness (i.e., the number of sinusoids
needed to describe the observations). The periodic Gaussian process
\citep{Mackay1998} is an alternative model for seasonal deformation
and is well suited for exploring the roughness of seasonal
deformation.  The periodic Gaussian process has zero mean and the
covariance function

\begin{equation}\label{eq:Periodic}
  T(t,t') = \phi^2 \exp\left(\frac{-\sin(\pi|t - t'|)^2}{2\tau^2}\right). 
\end{equation}

Realizations have annual periodicity and the roughness is controlled
by $\tau$. Decreasing $\tau$ has the same effect as including higher
frequency sinusoids in the seasonal model. The optimal value for
$\tau$ can be found with the REML method as described in Section
\ref{sec:NoiseModel}.

% issues and potential improvements on numerical consideratiosn

% Numerical considerations

% note that the compact covariance function is not a perfect solution
% because it is only compact if the signal has a short timescale. Long
% wavelength features will still have a dense covariance matrix. This is
% part of the reason why we omitted the temporal noise model.

There are numerical considerations as well ...
The transient strain rates estimated in this study are constrained by
about seven years of daily displacement observations from 94 GNSS
stations. It can be computationally intensive to evaluate eqs.
(\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) for a dataset
with this size. We significantly reduce the amount of memory needed to
estimate transient strain rates by describing the temporal covariance
of displacements with a compact Wendland covariance function. Using a
compact covariance function for our prior turns eqs.
(\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) into sparse
systems of equations, which we then solve with CHOLMOD. CHOLMOD is
designed for solving sparse, positive definite systems of equations.
The matrix being inverted in eqs. (\ref{eq:PosteriorMean3}) and
(\ref{eq:PosteriorCov3}) is not positive definite; however, we can use
another partitioned matrix inversion identity from \citet{Press2007}
to partition it into positive definite submatrices to be inverted.
Even when using a compact covariance function, it may still be
necessary to reduce the computational burden by dividing the data into
subsets and evaluating transient strain rates for each subset.

% additional applications

@@
